Exoplanet population inference and the abundance \\ of Earth analogs from noisy, incomplete catalogs

### 2014-08-23 document/ms.tex
note to self
re-writing inverse-detection-efficiency derivation

\end{itemize}
The first assumption---conditional independence of the candidates---is
reasonable since the dataset that we consider explicitly includes only single
transiting systems (\citealt{petigura}).
This is a limitation of the input catalog.
When considering other datasets it will be interesting and important to take
multiple transiting systems into account (\citealt{lissauer, tremaine, fang})
and this would be possible using conditional distributions (as developed by
\citealt{tremaine}).
The second assumption---neglecting false positives---is also strong and only
weakly justified by estimates of low false positive rates in the \kepler\ data
(\citealt{fressin-fp, morton}).
For this \paper, we will neglect this issue and only comment on the effects
but the prior distributions published by \citet{fressin-fp} could be directly
applied in a generalization of our method.

@@

The observed rate density \obs{\rate} is a quantitative description of the
rate density at which planets appear in the \citet{petigura} catalog; it is
not a description of the \True\ rate density of exoplanets.
Inasmuch as the detection efficiency $\completeness(\entry)$ is calculated
correctly, the function $\rate_\ratepars(\entry)$ will represent the \True\
rate density of exoplanets, at least where there is support in the data.
In practice, an estimate of the detection efficiency will not include every
decision or effect in the pipeline.
One example in this case is that the completeness function does not account
for the explicit removal of second (and third, and so on) planets.
In principle a more complete description of the detection efficiency can be
included and as this function becomes more accurate, our inferences about the
\True\ rate density $\rate_\ratepars(\entry)$ will be less biased.

@@

Finally, for all the results in this \paper, we'll model the rate density as
a piecewise constant step function
\begin{eqnarray}\eqlabel{rate-model}
\rate_\ratepars (\entry) &=& \left\{\begin{array}{ll}
\exp(\ratepar_1) & \entry \in \binarea_1,\\
\exp(\ratepar_2) & \entry \in \binarea_2,\\
\cdots \\
\exp(\ratepar_J) & \entry \in \binarea_J,\\
0 & \mathrm{otherwise}
\end{array}\right.
\end{eqnarray}
where the parameters $\ratepar_j$ are the log step heights and the bin edges
are fixed \foreign{a priori}.

@@

@@

In detail, any inference of the exoplanet population depends on the choices
made in the pipeline that creates the catalog.
Our results are based on a catalog that only includes single transiting
systems (\citealt{petigura}) so our conclusions---and theirs---must be
considered with this in mind.
In particular, the exclusion of second (and third, and so on) planets must
suppress the inferred rate density, especially at long periods and small
radii.

@@

\acknowledgments
We would like to thank Erik Petigura (Berkeley) for freely sharing his data
and code.
It is a pleasure to thank
Ruth Angus (Oxford),
Tom Barclay (NASA Ames),
Jo Bovy (IAS),
Eric Ford (PSU),
David Kipping (CfA),
Ben Montet (Caltech/Harvard), and
Scott Tremaine (IAS)
for helpful contributions to the ideas and code presented here.
We would also like to acknowledge the anonymous referee for suggestions that
substantially improved the paper.
This project was partially supported by the NSF (grant AST-0908357), NASA
(grant NNX08AJ48G), and the Moore--Sloan Data Science Environment at NYU.
This research builds on ideas generated at a three-week workshop supported by
NSF Grant DMS-1127914 to the Statistical and Applied Mathematical Sciences
Institute.
This research made use of the NASA \project{Astrophysics Data System}.

@@

It's now interesting to take a brief aside and discuss the connection between
our model and the commonly used inverse-detection-efficiency procedure
(\citealt{howard,dressing,petigura}).
This procedure involves making a weighted histogram of the catalog entries
where the weight for object $\entry_k$ is $1/\completeness(\entry_k)$.
This makes intuitive sense but the standard formulation does not have a clear
probabilistic justification or interpretation.

If we model the occurrence rate density as a histogram with $J$ fixed bin
volumes $\binarea_j$
\begin{eqnarray}
\rate_\ratepars (\entry) &=& \left\{\begin{array}{ll}
\exp(\ratepar_1) & \entry \in \binarea_1,\\
\exp(\ratepar_2) & \entry \in \binarea_2,\\
\cdots \\
\exp(\ratepar_J) & \entry \in \binarea_J,\\
0 & \mathrm{otherwise}
\end{array}\right.
\end{eqnarray}
then \eq{poisson-like} becomes
\begin{eqnarray}
\ln p(\{\entry_k\}\,|\,\ratepars) &=&
    \sum_{k=1}^K \sum_{j=1}^J \mathbf{1}[\entry_k \in
        \binarea_j]\,[\ln\completeness(\entry_k)+\ratepar_j]
    -\sum_{j=1}^J\exp(\ratepar_j)\,
        \int_{\binarea_j} \completeness(\entry)\dd\entry
\end{eqnarray}
where the indicator function $\mathbf{1}[\cdot]$ is one if $\cdot$ is true and
zero otherwise.
If we take the gradient of this function with respect to \ratepars\ and set it
equal to zero, we find the maximum likelihood result
\begin{eqnarray}\eqlabel{ml-ide}
\exp ({\ratepar_j}^*) &=&
\frac{N_j}{\int_{\binarea_j} \completeness(\entry)\dd\entry}
\end{eqnarray}
where $N_j$ is the number of objects that fall within the bin $j$.
We estimate the uncertainty $\delta\ratepar_j$ on this value by examining the
curvature of the log-likelihood function near the maximum and find
\begin{eqnarray}
\frac{\delta\exp({\ratepar_j}^*)}{\exp({\ratepar_j}^*)} &=&
    \frac{1}{\sqrt{N_j}}
\quad.
\end{eqnarray}

Since \eq{ml-ide} gives the maximum likelihood solution, this result is the
minimum-variance estimator for the rate density when the completeness function
$\completeness(\entry)$ is known and the catalog uncertainties are negligible.
This is \emph{almost} the same as the inverse-detection-efficiency procedure
and if the completeness function is constant in each bin \binarea, both
methods give the same result.
If the completeness function is not piecewise-constant on the same grid as the
rate density inference, the inverse-detection-efficiency procedure will
give a higher variance estimate of the occurrence rate density than
\eq{ml-ide}.
In this \paper, we don't use either of these inverse-detection-efficiency
procedures because they both fundamentally rely on the assumption of
\emph{negligible observational uncertainties}.
By applying \eq{money}, we derive the probabilistic generalization of this
model in the next section but the result is no longer analytic.

>>>>

\end{itemize}
The first assumption---conditional independence of the candidates---is
reasonable since the dataset that we consider explicitly includes only single
transiting systems (\citealt{petigura}).
The second assumption---neglecting false positives---is also strong and only
weakly justified by estimates of low false positive rates in the \kepler\ data
(\citealt{fressin-fp, morton}).
For this \paper, we will neglect this issue and only comment on the effects
but the prior distributions published by \citet{fressin-fp} could be directly
applied in a generalization of our method.

@@

The observed rate density \obs{\rate} is a quantitative description of the
rate density at which planets appear in the \citet{petigura} catalog; it is
not a description of the \True\ rate density of exoplanets.
Inasmuch as the detection efficiency $\completeness(\entry)$ is calculated
correctly, the function $\rate_\ratepars(\entry)$ will represent the \True\
rate density of exoplanets, at least where there is support in the data.
In practice, an estimate of the detection efficiency will not include every
decision or effect in the pipeline and as this function becomes more accurate,
our inferences about the \True\ rate density $\rate_\ratepars(\entry)$ will be
less biased.

@@

Finally, for all the results in this \paper, we'll model the rate density as
a piecewise constant step function
\begin{eqnarray}\eqlabel{rate-model}
\rate_\ratepars (\entry) &=& \left\{\begin{array}{ll}
\exp(\ratepar_1) & \entry \in \binarea_1,\\
\exp(\ratepar_2) & \entry \in \binarea_2,\\
\cdots \\
\exp(\ratepar_J) & \entry \in \binarea_J,\\
0 & \mathrm{otherwise}
\end{array}\right.
\end{eqnarray}
where the parameters $\ratepar_j$ are the log step heights and the bins
$\binarea_j$ are fixed \foreign{a priori}.

@@

\dfmtodo{Include Kipping's comments about eccentricity.}

@@

% In detail, any inference of the exoplanet population depends on the choices
% made in the pipeline that creates the catalog.
% Our results are based on a catalog that only includes single transiting
% systems (\citealt{petigura}) so our conclusions---and theirs---must be
% considered with this in mind.
% In particular, the exclusion of second (and third, and so on) planets must
% suppress the inferred rate density, especially at long periods and small
% radii.

@@

\acknowledgments
We would like to thank Erik Petigura (Berkeley) for freely sharing his data
and code.
It is a pleasure to thank
Ruth Angus (Oxford),
Tom Barclay (NASA Ames),
Jo Bovy (IAS),
Eric Ford (PSU),
David Kipping (CfA),
Ben Montet (Caltech/Harvard), and
Scott Tremaine (IAS)
for helpful contributions to the ideas and code presented here.
We would also like to acknowledge the anonymous referee and the Scientific
Editor, Eric Feigelson, for suggestions that
substantially improved the paper.
This project was partially supported by the NSF (grant AST-0908357), NASA
(grant NNX08AJ48G), and the Moore--Sloan Data Science Environment at NYU.
This research builds on ideas generated at a three-week workshop supported by
NSF Grant DMS-1127914 to the Statistical and Applied Mathematical Sciences
Institute.
This research made use of the NASA \project{Astrophysics Data System}.

@@

One huge benefit of the inverse-detection-efficiency procedure is its
simplicity.
Therefore, it's worth noting that there is a probabilistically justified
procedure that will always provide less biased results while being only
marginally more complicated.

The standard procedure involves making a weighted histogram of the catalog
entries where the weight for object $\entry_k$ is $1/\completeness(\entry_k)$.
This makes intuitive sense but it does not have a clear probabilistic
justification or interpretation.
As we will show below, the maximum likelihood result involves weighting the
points by the inverse of the \emph{integral} of the completeness function over
the bin area.

To motivate this derivation, let's start by considering the following
pathological example: a single bin where the completeness sharply drops from
one to zero halfway across the bin.
If we observe $K$ objects in this bin, we would have observed about $2K$
objects in a complete sample.
If we apply the inverse-detection-efficiency procedure to this dataset, each
sample will get unit weight because they are all found in the part of the bin
where the completeness is one.
Therefore, we would \emph{underestimate} the true rate in the bin by half.
It's clear in this specific case that giving the points a weight of two would
give a better solution and we'll derive the general result below.

If we model the occurrence rate density as a histogram with $J$ fixed bin
volumes $\binarea_j$ (\eqalt{rate-model}) then \eq{poisson-like} becomes
\begin{eqnarray}
\ln p(\{\entry_k\}\,|\,\ratepars) &=&
    \sum_{k=1}^K \sum_{j=1}^J \mathbf{1}[\entry_k \in
        \binarea_j]\,[\ln\completeness(\entry_k)+\ratepar_j]
    -\sum_{j=1}^J\exp(\ratepar_j)\,
        \int_{\binarea_j} \completeness(\entry)\dd\entry
\end{eqnarray}
where the indicator function $\mathbf{1}[\cdot]$ is one if $\cdot$ is true and
zero otherwise.
Taking the gradient of this function with respect to \ratepars\ and setting it
equal to zero, we find the maximum likelihood result
\begin{eqnarray}\eqlabel{ml-ide}
\exp ({\ratepar_j}^*) &=&
\frac{K_j}{\int_{\binarea_j} \completeness(\entry)\dd\entry}
\end{eqnarray}
where $K_j$ is the number of objects that fall within the bin $j$.
We estimate the uncertainty $\delta\ratepar_j$ on this value by examining the
curvature of the log-likelihood function near the maximum and find
\begin{eqnarray}
\frac{\delta\exp({\ratepar_j}^*)}{\exp({\ratepar_j}^*)}
&=& \frac{1}{\sqrt{K_j}}
\quad.
\end{eqnarray}

In our pathological example from above, the integral of the completeness
function over the bin is $1/2$, giving each sample the expected weight of $2$.
In more realistic cases, where the completeness function varies smoothly, the
inverse-detection-efficiency result will begin to agree with \eq{ml-ide} but
the severity of this bias will be very problem dependent.
Therefore, if you have a dataset with negligible observational uncertainties,
we recommend that you always apply \eq{ml-ide} instead of the standard
inverse-detection-efficiency procedure.
As the uncertainties become more significant, there is no longer an analytic
result and the method derived in this \paper\ is necessary.
